Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves

### 2025-07-25 ms/ms.qmd
added model summary tables

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We accounted for uncertainty in $\widehat{\mathrm{AA}}$ using the standard deviation of the posterior distribution of each $\widehat{\mathrm{AA}}$ estimate (@sec-aa-est). We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagnse*; LA0750 was added as sister to LA0716, two closely related populations of *S. pennellii*. The node separating LA0750 from LA0716 was placed half-way between the next deepest node. We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values. We selected the model with the lowest LOOIC value to generate posterior predictions for hypothesis testing.

@@

### MARKER - table of selected model 1

### MARKER - table of selected model 2

>>>>

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We accounted for uncertainty in $\widehat{\mathrm{AA}}$ using the standard deviation of the posterior distribution of each $\widehat{\mathrm{AA}}$ estimate (@sec-aa-est). We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. We used a robust regression approach by assuming $t$-distributed residuals, which makes estimates less susceptible to extreme values. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagnse*; LA0750 was added as sister to LA0716, two closely related populations of *S. pennellii*. The node separating LA0750 from LA0716 was placed half-way between the next deepest node. We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values. We selected the model with the lowest LOOIC value to generate posterior predictions for hypothesis testing.

@@

```{r fit_aa1, results = 'asis', echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
#| label: tbl-fit_aa1
#| tbl-cap: "Parameter estimates and 95% confidence intervals (CIs) from the posterior distribution of the selected model. Models potentially include fixed, random, and distributional effects of growth light intensity (shade and sun), measurement light intensity (low and high), and their interaction on amphistomy advantage ($\\mathrm{AA}$). The distributional effects refer to the effect of factors on residual variation in $\\mathrm{AA}$, denoted $\\sigma$, on a log-link scale. For phylogenetically structured random effects, we report the estimated standard deviation (SD)."

make_brms_summary_table = function(fit,
                                   ci_level = 0.95,
                                   digits = 2) {
  # Get posterior summary
  summ = posterior_summary(fit, prob = c((1 - ci_level) / 2, 1 - (1 - ci_level) / 2))
  summ_df = as.data.frame(summ)
  summ_df$param = rownames(summ_df)
  colnames(summ_df) = c("Estimate", "Est.Error", "CI_low", "CI_high", "Parameter")

  # Classify parameters
  summ_df = summ_df |>
    mutate(
      Type = case_when(
        grepl("^b_", Parameter) &
          !grepl("sigma|Intercept", Parameter) ~ "fixed effects",
        grepl("^b_Intercept$", Parameter) ~ "fixed effects",
        grepl("^b_", Parameter) &
          grepl("sigma", Parameter) ~ "distributional parameters",
        grepl("^sd_", Parameter) ~ "random effect SDs",
        TRUE ~ "Other"
      )
    )

  # Clean parameter names
  summ_df = summ_df |>
    mutate(
      Parameter = case_when(
        Parameter == "b_Intercept" ~ "$\\mathrm{AA}$ intercept (shade, low light)",
        Parameter == "b_light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$",
        Parameter == "b_light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$",
        Parameter == "b_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$",
        Parameter == "b_sigma_Intercept" ~ "$\\log \\sigma$ intercept (shade, low light)",
        Parameter == "b_sigma_light_treatmenthigh" ~ "effect of sun on $\\log \\sigma$",
        Parameter == "b_sigma_light_intensity2000" ~ "effect of high light on $\\log \\sigma$",
        Parameter == "sd_acc__Intercept" ~ "$\\mathrm{AA}$ among populations",
        Parameter == "sd_acc__light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$ among populations",
        Parameter == "sd_acc__light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$ among populations",
        Parameter == "sd_acc__light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$ among populations"
      )
    )

  # Select and format
  summ_df |>
    filter(Type != "Other") |>
    mutate(across(c(Estimate, CI_low, CI_high), ~ round(., digits))) |>
    mutate(`Estimate [95% CI]` = paste0(Estimate, " [", CI_low, ", ", CI_high, "]")) |>
    select(Type, Parameter, `Estimate [95% CI]`) |>
    arrange(match(
      Type,
      c(
        "fixed effects",
        "distributional parameters",
        "random effect SDs"
      )
    )) |>
    as_tibble()
}

summary_tbl = make_brms_summary_table(fit_aa1)

tbl_fixed = summary_tbl |> filter(Type == "fixed effects") |> select(-Type)
tbl_dist  = summary_tbl |> filter(Type == "distributional parameters") |> select(-Type)
tbl_rand  = summary_tbl |> filter(Type == "random effect SDs") |> select(-Type)

bind_rows(tbl_fixed, tbl_dist, tbl_rand) |>
  kable(
    col.names = c("Parameter", "Estimate [95\\% CI]"),
    booktabs = TRUE,
    format = "latex",
    escape = FALSE
  ) |>
  kable_styling(latex_options = c("striped")) |>
  pack_rows("Fixed effects",
            1,
            nrow(tbl_fixed),
            bold = FALSE,
            italic = TRUE) |>
  pack_rows(
    "Distributional parameters",
    nrow(tbl_fixed) + 1,
    nrow(tbl_fixed) + nrow(tbl_dist),
    bold = FALSE,
    italic = TRUE
  ) |>
  pack_rows(
    "Random effect SDs",
    nrow(tbl_fixed) + nrow(tbl_dist) + 1,
    nrow(tbl_fixed) + nrow(tbl_dist) + nrow(tbl_rand),
    bold = FALSE,
    italic = TRUE
  ) |>
  column_spec(1, width = "4in") |>
  column_spec(2, width = "2in")

```{r fit_aa2, results = 'asis', echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
#| label: tbl-fit_aa2
#| tbl-cap: "Parameter estimates and 95% confidence intervals (CIs) from the posterior distribution of the selected model with leaf mass per area (LMA) as potential mediator variable. Models potentially include fixed, random, and distributional effects of growth light intensity (shade and sun), measurement light intensity (low and high), $\\log \\mathrm{LMA}$, and their interactions on amphistomy advantage ($\\mathrm{AA}$). The distributional effects refer to the effect of factors on residual variation in $\\mathrm{AA}$, denoted $\\sigma$, on a log-link scale. LMA can also be affected by growth light intensity and population. For phylogenetically structured random effects, we report the estimated standard deviation (SD)."

make_brms_summary_table = function(fit,
                                   ci_level = 0.95,
                                   digits = 2) {
  # Get posterior summary
  summ = posterior_summary(fit, prob = c((1 - ci_level) / 2, 1 - (1 - ci_level) / 2))
  summ_df = as.data.frame(summ)
  summ_df$param = rownames(summ_df)
  colnames(summ_df) = c("Estimate", "Est.Error", "CI_low", "CI_high", "Parameter")

  # Classify parameters
  summ_df = summ_df |>
    mutate(
      Type = case_when(
        grepl("^b_", Parameter) &
          !grepl("sigma|Intercept", Parameter) ~ "fixed effects",
        grepl("^b_(aa|llma)_Intercept$", Parameter) ~ "fixed effects",
        grepl("^b_", Parameter) &
          grepl("sigma", Parameter) ~ "distributional parameters",
        grepl("^sd_", Parameter) ~ "random effect SDs",
        TRUE ~ "Other",
      ),
      resp = case_when(
        str_detect(Parameter, "_aa_") ~ "aa",
        str_detect(Parameter, "_llma_") ~ "lma",
        TRUE ~ NA_character_
      )
    )

  # Clean parameter names
  summ_df <- summ_df %>%
    mutate(
      Parameter = case_when(
        # AA
        Parameter == "b_aa_Intercept" ~ "$\\mathrm{AA}$ intercept (shade, low light)",
        Parameter == "b_aa_light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$",
        Parameter == "b_aa_light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$",
        Parameter == "b_aa_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$",
        Parameter == "b_sigma_aa_Intercept" ~ "$\\log \\sigma$ intercept (shade, low light)",
        Parameter == "b_sigma_aa_light_treatmenthigh" ~ "effect of sun on $\\log \\sigma$",
        Parameter == "b_sigma_aa_light_intensity2000" ~ "effect of high light on $\\log \\sigma$",
        Parameter == "sd_acc__aa_Intercept" ~ "$\\mathrm{AA}$ among populations",
        Parameter == "sd_acc__aa_light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$ among populations",
        Parameter == "sd_acc__aa_light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$ among populations",
        Parameter == "sd_acc__aa_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$ among populations",

        # LMA
        Parameter == "b_llma_Intercept" ~ "$\\log \\mathrm{LMA}$ intercept (shade)",
        Parameter == "b_llma_light_treatmenthigh" ~ "effect of sun on $\\log \\mathrm{LMA}$",
        Parameter == "b_sigma_llma_Intercept" ~ "$\\log \\sigma$ intercept (shade)",
        Parameter == "b_sigma_llma_light_treatmenthigh" ~ "effect of sun on $\\log \\sigma$",
        Parameter == "sd_acc__llma_Intercept" ~ "$\\log \\mathrm{LMA}$ among populations",
        Parameter == "sd_acc__llma_light_treatmenthigh" ~ "effect of sun on $\\log \\mathrm{LMA}$ among populations"
      )
    )

  # Select and format
  summ_df |>
    filter(Type != "Other") |>
    mutate(across(c(Estimate, CI_low, CI_high), ~ round(., digits))) |>
    mutate(`Estimate [95% CI]` = paste0(Estimate, " [", CI_low, ", ", CI_high, "]")) |>
    select(resp, Type, Parameter, `Estimate [95% CI]`) |>
    arrange(resp, match(
      Type,
      c(
        "fixed effects",
        "distributional parameters",
        "random effect SDs"
      )
    )) |>
    as_tibble()
}

summary_tbl = make_brms_summary_table(fit_aa2)

tbl_aa_fixed = summary_tbl |> filter(Type == "fixed effects", resp == "aa") |> select(-Type)
tbl_aa_dist  = summary_tbl |> filter(Type == "distributional parameters", resp == "aa") |> select(-Type)
tbl_aa_rand  = summary_tbl |> filter(Type == "random effect SDs", resp == "aa") |> select(-Type)
tbl_lma_fixed = summary_tbl |> filter(Type == "fixed effects", resp == "lma") |> select(-Type)
tbl_lma_rand  = summary_tbl |> filter(Type == "random effect SDs", resp == "lma") |> select(-Type)

bind_rows(tbl_aa_fixed,
          tbl_aa_dist,
          tbl_aa_rand,
          tbl_lma_fixed,
          tbl_lma_rand) |>
  select(-resp) |>
  kable(
    col.names = c("Parameter", "Estimate [95\\% CI]"),
    booktabs = TRUE,
    format = "latex",
    escape = FALSE
  ) |>
  kable_styling(latex_options = c("striped")) |>
  pack_rows(
    "Response: $\\mathrm{AA}$",
    1,
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand),
    escape = FALSE
  ) |>
  pack_rows(
    "~~Fixed effects",
    1,
    nrow(tbl_aa_fixed),
    escape = FALSE,
    bold = FALSE,
    italic = TRUE
  ) |>
  pack_rows(
    "~~Distributional parameters",
    nrow(tbl_aa_fixed) + 1,
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist),
    escape = FALSE,
    bold = FALSE,
    italic = TRUE
  ) |>
  pack_rows(
    "~~Random effect SDs",
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + 1,
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand),
    escape = FALSE,
    bold = FALSE,
    italic = TRUE
  ) |>
  pack_rows(
    "Response: $\\log~\\\\mathrm{LMA}$",
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + 1,
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + nrow(tbl_lma_rand),
    escape = FALSE
  ) |>
  pack_rows(
    "~~Fixed effects",
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + 1,
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed),
    escape = FALSE,
    bold = FALSE,
    italic = TRUE
  ) |>
  pack_rows(
    "~~Random effect SDs",
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + 1,
    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + nrow(tbl_lma_rand),
    escape = FALSE,
    bold = FALSE,
    italic = TRUE
  ) |>
  column_spec(1, width = "4.5in") |>
  column_spec(2, width = "2in")
